---
title: "R Notebook"
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

====================
Dataset Description
====================

The  dataset contains the hourly and daily count of rental bikes between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information. It was provided  on 2013-12-20 by Capital Bikeshare , a bicycle sharing system that serves Washington, D.C.

================================================================================
Attribute Information
================================================================================

Both hour.csv and day.csv have the following fields, except hr which is not available in day.csv

- instant: record index
- dteday : date
- season : season (1:springer, 2:summer, 3:fall, 4:winter)
- yr : year (0: 2011, 1:2012)
- mnth : month ( 1 to 12)
- holiday : weather day is holiday or not (extracted from [Web Link])
- weekday : day of the week
- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
+ weathersit : 
- 1: Clear, Few clouds, Partly cloudy, Partly cloudy
- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
- temp : Normalized temperature in Celsius. The values are derived via (t-t_min)/(t_max-t_min), t_min=-8, t_max=+39 (only in hourly scale)
- atemp: Normalized feeling temperature in Celsius. The values are derived via (t-t_min)/(t_max-t_min), t_min=-16, t_max=+50 (only in hourly scale)
- hum: Normalized humidity. The values are divided to 100 (max)
- windspeed: Normalized wind speed. The values are divided to 67 (max)
- casual: count of casual users
- registered: count of registered users
- cnt: count of total rental bikes including both casual and registered

===================================================================================
Questions to Explore
===================================================================================
1. when do we see the most casual riders? work days or weekends and holidays?
2. What days of the week were most rides taken?
3. During what seasons were most rides taken
4. During what weather conditions  were less rides taken and what was the average count during the season
5. Is there a corellation between holiday and total rental bikes
6. How does the weather affect the number of rides taken?
7. How does the temp affect the number of rides taken
8. Which variables are highly correlated with the total count of bike rentals/ Creating the best fit model
9. Can we make some decent predictions based on the model given the variables?
10. What are the  conclusions?


====================================================================
Research question/Hypothesis
======================================================================
Response variable : Rental Count(cnt)

H0:Rental count is not positively correlated with temperature
HA:Rental count is positively correlated with temperatrue




=============================================
Intsalling  and loading packages
=============================================
```{r}
# Setting global options
options("scipen" = 10)
options()$scipen

# Loading libraries
install.packages('dplyr')
install.packages('DMwR')
install.packages('reshape2')
install.packages('data.table')
install.packages('ggplot2')
library(dplyr)
library(DMwR)
library(reshape2)
library(data.table)
library(ggplot2)
library(RColorBrewer)

```



```{r}

# Importing the dataset and viewing structure
day=fread("https://raw.githubusercontent.com/FabienneIshimwe/Python-and-R/master/day.csv")
day=data.frame(day)
str(day)
```


#============================================
1. Data preprocessing
============================================

#1.1 Checking for null values

```{r}
is.null(day)
```


# Checking for outliers
```{r}

outliers=lofactor(day[,c(10:16)], k=5)
outliers=order(outliers, decreasing=T)[1:5]

n = nrow(day)
 pch = rep(".",n)
 pch[outliers] = "+"
col = rep("black", n)
col[outliers] = "red"
pairs(day[,c(10:16)], pch=pch, col=col)
 
```




#1.2 Adding new columns to facilitate analysis and the creation of better model

#Season
```{r}

day['season_new']=NA

day[day$season==1,]$season_new='spring'
day[day$season==2,]$season_new='summer'
day[day$season==3,]$season_new='fall'
day[day$season==4,]$season_new='winter'

day$season_new=factor(day$season_new)

```

#yr: year
```{r}

day['yr_new']=NA

day[day$yr==1,]$yr_new='2012'
day[day$yr==0,]$yr_new='2011'

day$yr_new=factor(day$yr_new)
```


#yr: weekday
```{r}
day['weekday_new']=NA
day[day$weekday==0,]$weekday_new='Sunday'
day[day$weekday==1,]$weekday_new='Monday'
day[day$weekday==2,]$weekday_new='Tuesday'
day[day$weekday==3,]$weekday_new='Wednesday'
day[day$weekday==4,]$weekday_new='Thursday'
day[day$weekday==5,]$weekday_new='Friday'
day[day$weekday==6,]$weekday_new='Saturday'

```

#workingday
```{r}
day['workingday_new']=NA

day[day$workingday==1,]$workingday_new='Regular Day'
day[day$workingday==0,]$workingday_new='Holiday or weekend'

day$workingday_new=factor(day$workingday_new)

```

#weathersit
```{r}

day['weathersit_new']=NA

day[day$weathersit==1,]$weathersit_new='good'
day[day$weathersit==2,]$weathersit_new='cloudy'
day[day$weathersit==3,]$weathersit_new='wet'
day$weathersit_new=factor(day$weathersit_new)

#day[day$weathersit==4,]$weathersit_new='bad'


```

#temp,atemp, hum, windspeed: Denormalizing the values using formula initialy used to normalise them.
```{r}

#- temp : Normalized temperature in Celsius. The values are derived via (t-t_min1)/(t_max1-t_min1), t_min1=-8, t_max1=+39 (only in hourly scale)

#- atemp: Normalized feeling temperature in Celsius. The values are derived via (t-t_min)/(t_max2-t_min2), t_min2=-16, t_max2=+50 (only in hourly scale

#- hum: Normalized humidity. The values are divided to 100 (max)

#- windspeed: Normalized wind speed. The values are divided to 67 (max)


t_max1=39
t_min1=-8
t_max2=50
t_min2=-16
  
#temp   
day['temp_new']=NA
day['temp_new']=day$temp*(t_max1-t_min1)+t_min1                                                                                                        
#atemp
day['atemp_new']=NA
day['atemp_new']=day$atemp*(t_max2-t_min2)+t_min2                                                                                                                                                                                                          
#hum                                                                                                    
day['hum_new']=NA
day['hum_new']= day$hum*100
                                                                                                    
#windspeed
day['windspeed_new']=NA
day['windspeed_new']= day$windspeed*67
                                                                                                    
                                                                                                  
```


# Chaning some of the columns to factor variables since they are categorical


```{r}

day$season=factor(day$season)
day$mnth=factor(day$mnth)
day$yr=factor(day$yr)
day$holiday=factor(day$holiday)
day$weekday=factor(day$weekday)
day$workingday=factor(day$workingday)
day$weathersit=factor(day$weathersit)

    
```

# Changing dteday to a date column
```{r}
day=day %>%
   mutate(dteday=as.Date(dteday, format = "%Y-%m-%d"))
```

```{r}
day$mnth_new <- factor(months(day$dteday)
                      ,levels = c("January"
                                  ,"February"
                                  ,"March"
                                  ,"April"
                                  ,"May"
                                  ,"June"
                                  ,"July"
                                  ,"August"
                                  ,"September"
                                  ,"October"
                                  ,"November"
                                  ,"December")
                      ,ordered = TRUE)



```



# View the new structure and sample of the dataset
```{r}
str(day)

head(day)
```


#====================================
#2. EDA       
#=====================================


# correlation between numerical variabels

```{r}
hm.palette <- colorRampPalette(rev(brewer.pal(11, 'BrBG')), space='Lab')

dayCor <- cor(day[, c(10:16)], use="complete.obs")
dayCorMelt <- reshape2::melt(dayCor, varnames=c("x", "y"), value.name="Correlation")
dayCorMelt <- dayCorMelt[order(dayCorMelt$Correlation),]
ggplot(dayCorMelt, aes(x,y, fill = Correlation)) + geom_tile()+
  scale_fill_gradientn(colours = hm.palette(100))
```


# Distribution of the response variable (cnt)  as well as redistered and cusual renters
```{r}

# cnt
ggplot(day, aes(cnt) )+
  geom_histogram(bins  = 80, fill='tan4', color='yellowgreen')+
geom_vline(aes(xintercept=mean(cnt, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)
# casual
ggplot(day, aes(casual))+
  geom_histogram(bins  = 80, fill='tan2', color='yellowgreen')+
geom_vline(aes(xintercept=mean(casual, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)
# registered
ggplot(day, aes(registered) )+
  geom_histogram(bins  = 80, fill='wheat4', color='yellowgreen')+
geom_vline(aes(xintercept=mean(registered, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)
mean(day$casual)

```


Distribution of numerical explanatory variables
```{r}

#Actual Temperature
ggplot(day, aes(temp_new)) +
  geom_histogram(bins  = 80, fill='turquoise4', color='skyblue')+
geom_vline(aes(xintercept=mean(temp_new, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)+
  labs(y = "Total rental count", x='Actual Temperature',title='Actual Temperature Dististribution')


# Apparent temperature
ggplot(day, aes(atemp) )+
  geom_histogram(bins  = 80, fill='tan4', color='skyblue')+
geom_vline(aes(xintercept=mean(atemp, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)+
labs( x='Aapprent Temperature',title='Apparent Temperature Dististribution')

# Humidity
ggplot(day, aes(hum_new) )+
  geom_histogram(bins  = 80, fill='tan2', color='yellowgreen')+
geom_vline(aes(xintercept=mean(hum_new, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)+
  labs(x='Humidity',title='Humidity Dsitribution')

# windispeed
ggplot(day, aes(windspeed_new) )+
  geom_histogram(bins  = 80, fill='darkorchid4', color='yellowgreen')+
geom_vline(aes(xintercept=mean(windspeed_new, na.rm=T)),   # Ignore NA values for mean
               color="red", linetype="dashed", size=2)+
  labs( x='Windspeed',title='Windspeed Distribution')

```


# dsitribution of categorical variables

```{r}
# year Total count
ggplot(day, aes(x=yr_new,y=cnt, fill=yr_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='rental count distribution over  2 years', x='year')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

# year casual count
ggplot(day, aes(x=yr_new,y=casual, fill=yr_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='casual count distribution over  2 years', x='year')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
    coord_cartesian(ylim = c(0, 10000))


# weekday total count

ggplot(day, aes(x=weekday_new,y=cnt, fill=weekday_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='rental count distribution vs weekday', x='Weekday')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

# weekday casual count

ggplot(day, aes(x=weekday_new,y=casual, fill=weekday_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='casual count distribution vs weekday', x='Weekday')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
    coord_cartesian(ylim = c(0, 10000))


#  seasons total count
ggplot(day, aes(x=season_new,y=cnt, fill=season_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='rental count distribution vs season', x='Season')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

#  seasons casual count
ggplot(day, aes(x=season_new,y=casual, fill=season_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='casual count distribution vs season', x='Season')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
    coord_cartesian(ylim = c(0, 10000))


# weathersit total count
ggplot(day, aes(x=weathersit_new,y=cnt, fill=weathersit_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='rental count distribution vs weather condition', x='Weather condition')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

# weathersit casual count
ggplot(day, aes(x=weathersit_new,y=casual, fill=weathersit_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='casual count distribution vs weather condition', x='Weather condition')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
    coord_cartesian(ylim = c(0, 10000))


# month total count
ggplot(day, aes(x=mnth_new,y=cnt, fill=mnth_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='rental count distribution over months', x='month')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

# month casual count
ggplot(day, aes(x=mnth_new,y=casual, fill=mnth_new))+
    geom_boxplot()+
     labs(y = "Total rental count ", title='casual count distribution over months', x='month')+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
  coord_cartesian(ylim = c(0, 10000))


```


#Percentage of registered bikers on work days and non work days

```{r}

day['casualPro']=day$casual/day$cnt*100
day['registeredPro']=100-day$casualPro

ggplot() + geom_line(aes(y = registeredPro, x = dteday, color=workingday_new),
                           data = day)+
    labs(x="Year", y="Casual riders total percentage(%)") +
  ggtitle("Percentage of casual bikers on work days and non work days")+
  
  theme(axis.line = element_line(size=1, colour = "black"), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), panel.border = element_blank(),
        panel.background = element_blank()
        )
```







# Simple linear regression Analysis with numerical variables


```{r}
ggplot(day, aes(x=temp_new, y=cnt)) +
    geom_point(shape=1) +   
    geom_smooth(method=lm)  





ggplot(day, aes(x=atemp_new, y=cnt)) +
    geom_point(shape=1) +    
    geom_smooth(method=lm)   




ggplot(day, aes(x=hum_new, y=cnt)) +
    geom_point(shape=1) +    
    geom_smooth(method=lm)   




ggplot(day, aes(x=windspeed_new, y=cnt)) +
    geom_point(shape=1) +    
    geom_smooth(method=lm)   

lm1=lm(cnt ~ atemp_new, data = day)
lm2=lm(cnt ~ hum_new, data = day)
lm3=lm(cnt ~ temp_new, data = day)
lm4=lm(cnt ~ windspeed_new, data = day)


lm1
lm2
lm3
lm4
```

# simple linear regression model evaluation
```{r}

plot(lm1, col = "green")

plot(lm2, col = "blue")

plot(lm3, col = "yellow")

plot(lm4, col = "brown")
```





# Multilinear regression

```{r}

# splitting data into train and test

## 75% of the sample size
smp_size =floor(0.75 * nrow(day))

set.seed(123)
train_ind =sample(seq_len(nrow(day)), size = smp_size)

train= day[train_ind, ]
test =day[-train_ind, ]

```

# removing variables that with p-values equal to NA 
```{r}


# All predictors included except dteday and yr
model <- lm(cnt~+season+mnth+holiday+weekday+workingday+weathersit+temp+atemp+hum+windspeed,data=train)
summary(model)
```

```{r}
#removing the ones with p-values less >0.005

#1. remove mnth
model <- lm(cnt~+season+holiday+weekday+workingday+weathersit+temp+atemp+hum+windspeed,data=train)
summary(model)


#2. remove workingday
model <- lm(cnt~+season+holiday+weekday+weathersit+temp+atemp+hum+windspeed,data=train)
summary(model)

#3. remove weekday
model <- lm(cnt~+season+holiday+weathersit+temp+atemp+hum+windspeed,data=train)
summary(model)
#4. remove atemp
model <- lm(cnt~+season+holiday+weathersit+temp+hum+windspeed,data=train)
summary(model)
#5. remove weathersit
model <- lm(cnt~+season+holiday+temp+hum+windspeed,data=train)
summary(model)

#5. remove holiday
model <- lm(cnt~+season+temp+hum+windspeed,data=train)
summary(model)


summary(model)
```
# predicting

```{r}
p=predict(model, test[1:5,])
p

head(test)

plot(model,col='green')
```




